---
title: "ebird"
output: html_document
date: "2026-02-27"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(janitor)
library(naniar)
library(ggmap)
library(leaflet)
library(patchwork)
library(ggdensity)
library(paletteer)
library(RColorBrewer)
```

```{r}
getwd()
```
## Prepping the data: Reading into R, establishing NA's, selecting for relevent information, and writing a new file.

```{r}
#species <- read.delim("ebd_US-CA-113_201801_202312_smp_relJan-2026.txt", quote = "", na="") %>% clean_names()
```

```{r}
#species <- species %>%
#  select(taxonomic_order, common_name, scientific_name, observation_count, locality, latitude, longitude, observation_date, observer_id, sampling_event_identifier, observation_type, duration_minutes, effort_distance_km, number_observers)
```

```{r}
#species <- species %>%
#  separate(observation_date, into = c("year", "month", "day"), sep = "-")
```

```{r}
#species <- species %>%
#  mutate(across(c("year", "month", "day", "observation_type"), as.factor))
```

```{r}
#species <- species %>%
 # filter(year!=2018 & year!= 2022 & year!=2023 & observation_count!="X")
```

```{r}
#species <- species %>%
# filter(observation_type!="Historical" & observation_type!="Incidental" & observation_type!="Nocturnal Flight Call Count" & observation_type!="Banding")
```

```{r}
#species <- species %>%
#  mutate(observation_count=na_if(observation_count, "X"))
```

```{r}
#species %>%
#  miss_var_summary()
```

```{r}
#write_csv(species, "species_data.csv")
```

```{r}
#samples <- read.delim("ebd_US-CA-113_201801_202312_smp_relJan-2026_sampling.txt", quote = "", na="") %>% clean_names()
```

```{r}
#total_data <- full_join(species, samples, by="sampling_event_identifier")
```

```{r}
#total_data <- total_data %>%
#  select(common_name, scientific_name, observation_count, locality.y, latitude.y, longitude.y, observer_id.y, sampling_event_identifier, observation_type.y, duration_minutes.y, effort_distance_km.y, number_observers.y, year.y, month.y, day.y)
```

```{r}
#glimpse(total_data)
```

```{r}
#write_csv(total_data, "total_data.csv")
```
```{r}
#samples <- samples %>%
#  select(locality, latitude, longitude, observer_id, sampling_event_identifier, observation_type, duration_minutes, effort_distance_km, number_observers, observation_date) %>%
#  separate(observation_date, into = c("year", "month", "day"), sep = "-") %>%
#  mutate(across(c("year", "month", "day"), as.factor))
```

```{r}
#samples <- samples %>%
#  mutate(across(c("year", "month", "day", "observation_type"), as.factor)) %>%
#  filter(observation_type!="Historical" & observation_type!="Incidental" & observation_type!="Nocturnal Flight Call Count" & observation_type!="Banding")
```

```{r}
#write_csv(samples, "sample_data.csv")
```

```{r message=FALSE, warning=FALSE}
library(ggmap)
#library(patchwork)
library(ggdensity)
#library(paletteer)
library(RColorBrewer)
library(shiny)
library(shinydashboard)
```

```{r}
samples <- read_csv("sample_data.csv") %>% clean_names()
species <- read_csv("species_data.csv") %>% clean_names()
total_data <- read_csv("total_data.csv") %>% clean_names()
```

## Registering and creating the maps.

```{r}
register_stadiamaps("c7554538-a17b-488f-9eb4-62494d62e9e5", write = FALSE)
```

#Creating the Yolo County map:

```{r}
samples %>%
  select(latitude, longitude) %>%
  summary()
```

```{r}
lat_yolo <- c(38.32, 38.93)
long_yolo <- c(-122.4, -121.5)
bbox_yolo <- make_bbox(long_yolo, lat_yolo, f=0.03)
```

```{r}
map_yolo <- get_stadiamap(bbox_yolo, maptype = "stamen_terrain", zoom = 11)
```

```{r}
ggmap(map_yolo)+
  labs(title = "Yolo County", x="Longitude", y="Latitude")+
  theme(
    axis.title.x = element_text(vjust = -1), # Negative vjust moves it down/outward
    axis.title.y = element_text(vjust = 3)
  )
```

```{r}
ggmap(map_yolo) +
  geom_point(data=samples, aes(longitude, latitude), size=0.3, color="black", alpha=0.8)+
  labs(title = "Checklist Locations for 2018-2023 for Yolo County", x="Longitude", y="Latitude")+
  theme(
    axis.title.x = element_text(vjust = -1), # Negative vjust moves it down/outward
    axis.title.y = element_text(vjust = 3)  # Positive vjust moves it up/outward
  )
```

```{r}
ggmap(map_yolo) +
  geom_density_2d_filled(data=samples, aes(longitude, latitude), size=1, alpha=0.5, bins=5)+
  labs(title = "Density of Checklists for 2018-2023 in Yolo County", x="Longitude", y="Latitude")+
  theme(
    axis.title.x = element_text(vjust = -1), # Negative vjust moves it down/outward
    axis.title.y = element_text(vjust = 3)  # Positive vjust moves it up/outward
  )
```
```{r}
dens_check <- samples %>%
  filter(year==2018 | year==2020 | year==2023)
```

```{r}
ggmap(map_yolo) +
  geom_density_2d_filled(data=dens_check, aes(longitude, latitude), size=1, alpha=0.5)+
  labs(title = "Density of Checklists for 2018, 2020, and 2023 in Yolo", x="Longitude", y="Latitude")+
  theme(axis.title.x = element_text(vjust = -1),
        axis.title.y = element_text(vjust = 2),
        axis.text.x=element_text(angle=60, hjust = 1))+
  facet_wrap(~ year)
```

#Creating the Davis map:

```{r}
lat_davis <- c(38.525, 38.575)
long_davis <- c(-121.678, -121.792)
bbox_davis <- make_bbox(long_davis, lat_davis, f=0.03)
```

```{r}
map_davis <- get_stadiamap(bbox_davis, maptype = "stamen_terrain", zoom = 14)
```

```{r}
ggmap(map_davis)
```

```{r}
ggmap(map_davis) +
  geom_point(data=samples, aes(longitude, latitude), size=0.9, alpha=0.8)+
    labs(title = "Checklist Locations for 2018-2023 for Davis", x="Longitude", y="Latitude")+
  theme(axis.title.x = element_text(vjust = -1), axis.title.y = element_text(vjust = 2))+
  scale_color_brewer(palette = "Dark2")
```

```{r}
ggmap(map_davis) +
  geom_density_2d_filled(data=samples, aes(longitude, latitude), size=1, alpha=0.5)+
  labs(title = "Density of Checklists for 2018-2023 in Davis", x="Longitude", y="Latitude")+
  theme(axis.title.x = element_text(vjust = -1), axis.title.y = element_text(vjust = 2))
```

```{r}
ggmap(map_davis) +
  geom_density_2d_filled(data=dens_check, aes(longitude, latitude), size=1, alpha=0.5)+
  labs(title = "Density of Checklists for 2018, 2020, and 2023 in Davis", x="Longitude", y="Latitude")+
  theme(axis.title.x = element_text(vjust = -1),
        axis.title.y = element_text(vjust = 2),
        axis.text.x=element_text(angle=60, hjust = 1))+
  facet_wrap(~ year)
```

##Prepping the seasonal data


```{r}
season_observed <- samples %>%
  mutate(across(c("month"), as.numeric)) %>%
  mutate(season=case_when(month>=3 & month<6 ~ "spring",
                            month>=6 & month<9  ~"summer",
                            month>=9 & month<12 ~ "fall",
                            month>=12 | month<3  ~ "winter")) %>%
  mutate(season=factor(season,
                                 levels=c("spring", "summer", "fall", "winter")))
```

```{r}
season_observed %>%
  filter(season=="spring")%>%
  group_by(locality) %>%
  count(sort=T)
```

```{r}
season_observed_local <- season_observed %>%
  select(locality, season)
```

```{r}
season_obs_w <- season_observed_local %>%
  count(season, locality, sort=T) %>%
  pivot_wider(names_from = season,
              values_from = n)
```

```{r}
season_obs_w %>%
  arrange(desc(spring)) %>%
  slice_head(n=5) %>%
  ggplot(aes(x=reorder(locality, spring), y=spring, fill=locality))+
  geom_col()+
  coord_flip()+
  scale_fill_brewer(palette = "Set3")+
  labs(title = "Top 5 Birding Locations in Spring", x="Location", y= "Number of Observations")+
  theme(legend.position = "none")
```
```{r}
samples %>%
  filter(latitude>=38.525
         & latitude<=38.575 
         & longitude<=-121.678 
         & longitude>=-121.792) %>%
  summarise(total = n())
```
```{r}
samples %>%
  count()
```

```{r}
season_obs_w %>%
  arrange(desc(summer)) %>%
  slice_head(n=5) %>%
  ggplot(aes(x=reorder(locality, summer), y=summer, fill=locality))+
  geom_col()+
  coord_flip()+
  scale_fill_brewer(palette = "Set3")+
  labs(title = "Top 5 Birding Locations in Summer", x="Location", y= "Number of Observations")+
  theme(legend.position = "none")
```

```{r}
season_obs_w %>%
  arrange(desc(fall)) %>%
  slice_head(n=5) %>%
  ggplot(aes(x=reorder(locality, fall), y=fall, fill=locality))+
  geom_col()+
  coord_flip()+
  scale_fill_brewer(palette = "Set3")+
  labs(title = "Top 5 Birding Locations in Fall", x="Location", y= "Number of Observations")+
  theme(legend.position = "none")
```

```{r}
season_obs_w %>%
  arrange(desc(winter)) %>%
  slice_head(n=5) %>%
  ggplot(aes(x=reorder(locality, winter), y=winter, fill=locality))+
  geom_col()+
  coord_flip()+
  scale_fill_brewer(palette = "Set3")+
  labs(title = "Top 5 Birding Locations in Winter", x="Location", y= "Number of Observations")+
  theme(legend.position = "none")
```

```{r}
season_observed %>%
  mutate(across(c("year"), as.factor)) %>%
  ggplot(aes(x=season, fill=year))+
  geom_bar(position = "dodge")+
  scale_fill_brewer(palette = "Set3")+
  labs(title = "Checklist Count Per Year By Season for Yolo County", x="Season", y="Checklist Count", fill="Year")
```

## Top 5 Birding Locations in Yolo County

```{r}
samples %>%
  count(locality, sort=T)
```

```{r}
samples %>%
  count(locality, sort=T) %>%
  slice_head(n=5) %>%
  ggplot(aes(x= reorder(locality,n), y=n, fill=locality))+
  geom_col()+
  coord_flip()+
  scale_fill_brewer(palette = "Set3")+
  labs(title = "Top 5 Birding Locations Yolo County", x="Location", y= "Number of Observations")+
  theme(legend.position = "none")
```
## Checklist Counts Per Year in Yolo County

```{r}
samples %>%
  group_by(year) %>%
  count() 
```
```{r}
samples %>%
  mutate(across(c("year"), as.factor)) %>%
  group_by(year) %>%
  ggplot(aes(x=year, fill=year))+
  geom_bar()+
  scale_fill_brewer(palette = "Set3")+
  labs(title = "Amount of Checklists per Year from 2018-2023", x="Year", y= "Number of Checklists", fill="Year")+
  geom_text(aes(label = after_stat(count)), stat = "count", vjust = -0.5, size = 3.5)
```


```{r}
samples %>%
  mutate(across(c("year"), as.factor))%>%
  group_by(year, month) %>%
  count() %>%
  ggplot(aes(x=month, y=n, group = year, color=year))+
  geom_line()+
  scale_color_brewer(palette = "Dark2")+
  labs(title = "Amount of Checklists per Month from 2018-2023", x="Month", y= "Number of Checklists", color="Year")
```

```{r}

```


## Are more people sitting or traveling?

```{r}
samples %>%
  group_by(observation_type) %>%
  count(sort=T)
```

```{r}
season_observed %>%
  filter(observation_type!="Area") %>%
  mutate(across(c("year"), as.factor)) %>%
  group_by(season, observation_type, year) %>%
  count() %>%
  ggplot(aes(x=year, y=n, fill=year))+
  geom_col()+
  facet_wrap(season ~ observation_type, nrow=2)
```

```{r}
samples %>%
  filter(observation_type!="Area") %>%
  mutate(across(c("year"), as.factor)) %>%
  group_by(observation_type, year) %>%
  count() %>%
  ggplot(aes(x=observation_type, y=n, fill=observation_type))+
  geom_col()+
  facet_wrap(~ year, nrow=2)
```
##

```{r}
species %>%
  group_by(year) %>%
  distinct(common_name) %>%
  count()
```
```{r}
species %>%
  mutate(across(c("year"), as.factor)) %>%
  group_by(year) %>%
  distinct(common_name) %>%
  count() %>%
  ggplot(aes(x=year, y=n, fill = year))+
  geom_col()+
  scale_fill_brewer(palette = "Set3")+
  labs(title = "Number of Distinct Species Seen by Year", x="Year", y= "Number of Species", fill="Year")+
  geom_text(aes(label = n), vjust = 1.5, size = 3.5)
```


```{r}
samples %>%
  mutate(across(c("year"), as.factor)) %>%
  group_by(year) %>%
  count()
```

```{r}
species %>%
  distinct(common_name) %>%
  count()
```
```{r}
species %>%
  group_by(year) %>%
  count() %>%
  arrange(desc(n))
```
```{r}
species %>%
  mutate(across(c("year"), as.factor)) %>%
  group_by(year) %>%
  count() %>%
  arrange(desc(n)) %>%
  ggplot(aes(x=year, y=n, fill=year))+
  geom_col()+
  labs(title = "Number of Total Birds Seen by Year", x="Year", y= "Number of Birds", fill="Year")+
  geom_text(aes(label = n), vjust = 1.5, size = 3.5)+
  scale_fill_brewer(palette = "Set3")
```


```{r}
samples %>%
  mutate(across(c("year"), as.factor)) %>%
  drop_na(duration_minutes) %>%
  ggplot(aes(x=year, y=log10(duration_minutes), fill=year))+
  geom_boxplot()+
  scale_fill_brewer(palette = "Set3")
```
```{r}
samples %>%
  mutate(across(c("year"), as.factor)) %>%
  filter(year==2018 | year==2020 | year==2023) %>%
  drop_na(duration_minutes) %>%
  ggplot(aes(x=log10(duration_minutes), fill=year))+
  geom_density(alpha=0.3)+
  scale_fill_brewer(palette = "Set3")
```

```{r}
samples %>%
  mutate(across(c("year"), as.factor)) %>%
  drop_na(effort_distance_km) %>%
  ggplot(aes(x=year, y=log10(effort_distance_km), fill=year, na.rm=T))+
  geom_boxplot()+
  scale_fill_brewer(palette = "Set3")
```
```{r}
samples %>%
  mutate(across(c("year"), as.factor)) %>%
  filter(year==2018 | year==2020 | year==2023) %>%
  drop_na(effort_distance_km) %>%
  ggplot(aes(x=log10(effort_distance_km), fill=year, na.rm=T))+
  geom_density(alpha=0.3)+
  scale_fill_brewer(palette = "Set3")
```

```{r}
samples %>%
  mutate(across(c("year"), as.factor)) %>%
  filter(observation_type=="Traveling" & (year==2018 | year==2020 | year==2023)) %>%
  drop_na(effort_distance_km, duration_minutes)%>%
  ggplot(aes(x=log10(effort_distance_km), y=log10(duration_minutes), color=year))+
  geom_point(size=0.5)+
  geom_smooth(method=lm, se = F)+
  scale_color_brewer(palette = "Dark2")
```

```{r}
obs_total <- species %>%
  group_by(sampling_event_identifier) %>%
  summarise(observation_count_total=sum(observation_count))
```
```{r}
dur <- species %>%
  select(duration_minutes, sampling_event_identifier, year)%>%
  distinct(sampling_event_identifier, duration_minutes, year)
```
```{r}
dur_vs_obs <- full_join(obs_total, dur, by="sampling_event_identifier")
```

```{r}
dur_vs_obs %>%
  mutate(across(c("year"), as.factor)) %>%
  drop_na(observation_count_total, duration_minutes) %>%
  ggplot(aes(x=log10(duration_minutes), y=log10(observation_count_total), color=year))+
  geom_point(size=0.1)+
  geom_smooth(method=lm, se=F)+
  scale_color_brewer(palette = "Dark2")
```
```{r}
effort <- species %>%
  select(effort_distance_km, sampling_event_identifier, year)%>%
  distinct(sampling_event_identifier, effort_distance_km, year)
```

```{r}
effort_obs <- full_join(effort, obs_total, by="sampling_event_identifier")
```

```{r}
effort_obs %>%
  mutate(across(c("year"), as.factor)) %>%
  drop_na(observation_count_total,effort_distance_km) %>%
  ggplot(aes(x=log10(effort_distance_km), y=log10(observation_count_total), color=year))+
  geom_point(size=0.5)+
  geom_smooth(method=lm, se=F)+
  scale_color_brewer(palette = "Dark2")
```

```{r}
people <- species %>%
  select(number_observers, sampling_event_identifier, year)%>%
  distinct(sampling_event_identifier, number_observers, year)
```

```{r}
people_obs <- full_join(people, obs_total, by="sampling_event_identifier")
```

```{r}
people_obs %>%
  mutate(across(c("year"), as.factor)) %>%
  filter(sampling_event_identifier!="S93953515") %>%
  drop_na(number_observers) %>%
  ggplot(aes(x=number_observers, y=log10(observation_count_total), color=year))+
  geom_point(size=1)+
  geom_smooth(method=lm, se=F)+
  scale_color_brewer(palette = "Dark2")
```


##Shiny App Map Portion

```{r}
seasons <- read_csv("total_data_seasons.csv")
```
```{r}
seasons <- seasons %>%
       mutate(across(c("year_y", "season"), as.factor))
```

```{r}

ui <- dashboardPage(
  
  dashboardHeader(title="Map of Yolo County and Davis by Season and Year", titleWidth = 500),
  
  dashboardSidebar(disable=T),
  
  dashboardBody(
  
  selectInput("x",
               "Select Year",
               choices=c("2018",
                         "2019",
                         "2020",
                         "2021",
                         "2022",
                         "2023"),
               selected="2018"),

  selectInput("y",
              "Select Season",
              choices=c("Winter",
                        "Spring",
                        "Summer",
                        "Fall"),
              selected="Winter"),
  
  fluidRow(
    # Left Map (width 6 out of 12 columns)
    column(6,
      h3("Map 1: Yolo County"),
      plotOutput("map1")
    ),

    # Right Map (width 6 out of 12 columns)
    column(6,
      h3("Map 2: Davis"),
      plotOutput("map2")
    )
)))





server <- function(input, output, session) {

  
  
    output$map1 <- renderPlot({
      
      dynamic_title <- paste("Density of Checklists in", input$y, "of", input$x)
      
     filtered_selection <- seasons %>%
        filter(year_y==input$x & season==input$y) 
    
      ggmap(map_yolo) +
  geom_density_2d_filled(data=filtered_selection, aes(longitude_y, latitude_y), size=1, alpha=0.5)+
  labs(title = dynamic_title, x="Longitude", y="Latitude")+
  theme(
    axis.title.x = element_text(vjust = -1), # Negative vjust moves it down/outward
    axis.title.y = element_text(vjust = 3)  # Positive vjust moves it up/outward
  )})
     
      output$map2 <- renderPlot({
        
        dynamic_title <- paste("Density of Checklists in", input$y, "of", input$x)
        
        filtered_selection <- seasons %>%
        filter(year_y==input$x & season==input$y)
      
      ggmap(map_davis) +
  geom_density_2d_filled(data=filtered_selection, aes(longitude_y, latitude_y), size=1, alpha=0.5)+
  labs(title = dynamic_title, x="Longitude", y="Latitude")+
  theme(
    axis.title.x = element_text(vjust = -1), # Negative vjust moves it down/outward
    axis.title.y = element_text(vjust = 3)  # Positive vjust moves it up/outward
  ) 
      
    
  })
  
  
  
}





shinyApp(ui, server)
```

```{r}
samples_time <- read.delim("ebd_US-CA-113_201801_202312_smp_relJan-2026_sampling.txt", quote = "", na="") %>% clean_names()
```

```{r}
samples_time <- samples_time %>%
  select(locality, latitude, longitude, observer_id, sampling_event_identifier, observation_type, duration_minutes, effort_distance_km, number_observers, observation_date, time_observations_started) %>%
  separate(observation_date, into = c("year", "month", "day"), sep = "-") %>%
  separate(time_observations_started, into = c("hour", "minute", "sec"), sep = ":") %>%
  mutate(across(c("year", "month", "day"), as.factor))
```

```{r}
samples_time %>%
  drop_na(hour) %>%
  filter(year==2018 | year==2020 | year==2023) %>%
  group_by(hour, year, month) %>%
  count() %>%
  ggplot(aes(x=hour, y=n, group = year, color=year))+
  geom_line()+
  facet_wrap(~month)
```

```{r}
samples_time %>%
  drop_na(hour)  %>%
  group_by(hour, month) %>%
  count() %>%
  ggplot(aes(x=hour, y=n, group = month, color=month))+
  geom_line()
```



























