---
title: "Josephine_WIP"
author: "Josephine Wright"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(janitor)
library(RColorBrewer)
library(ggplot2)
library(paletteer)
library(patchwork)
library(naniar)
```

```{r}
setwd("~/Desktop/BIS15L_W2026_Group-8/rbird")
```

Data
```{r}
sample_data <- read_csv("sample_data.csv") %>%
  clean_names()
```
```{r}
species_data <- read_csv("species_data.csv") %>%
  clean_names()
```
```{r}
total_data <- read_csv("total_data.csv") %>%
  clean_names()
```

Removing data from total_data where the only information is sampling_event_identifier
```{r}
#total_data <- total_data %>%
#  filter(!is.na(common_name))
```

Adding season column
```{r}
total_data <- total_data %>%
  mutate(across(c("month_y"), as.numeric)) %>%
  mutate(season = case_when(
    month_y %in% c(12, 1, 2) ~ "Winter",
    month_y %in% c(3, 4, 5) ~ "Spring",
    month_y %in% c(6, 7, 8) ~ "Summer",
    month_y %in% c(9, 10, 11) ~ "Fall",
    TRUE ~ NA_character_)) %>% 
  mutate(season = factor(season, levels = c("Winter", "Spring", "Summer", "Fall")))
total_data
```

**General Data Analyses Pertaining to Birds**

How many bird species are represented by the data?
```{r}
total_data %>% 
  distinct(common_name) %>% 
  count()
```

How many birds were seen but unidentified?
```{r}
total_data %>% 
  filter(is.na(common_name)) %>% 
  count()
```

```{r}
total_data %>% 
  select(common_name) %>%
  miss_var_summary()
```
How many birds total are represented by the data?
```{r}
total_data %>% 
  count()
```

**Top Five Bird Observations**

Top five bird observations between 2019 and 2021 (total)
```{r}
top_five_total <- total_data %>%
  filter(year_y %in% c(2019, 2020, 2021)) %>% 
  group_by(common_name) %>%
  summarise(count = n()) %>%
  arrange(desc(count)) %>%
  slice_head(n = 5)
top_five_total
```

```{r}
top_five_total %>% 
ggplot(aes(x = common_name, y = count, fill = common_name)) +
  geom_bar(stat = "identity") +
  coord_flip()+
  scale_fill_brewer(palette = "Set3")+
  labs(title = "Top Five Birds with Most Observations",
       x = "Common Name",
       y = "Count of Observations",
       fill="Common Name") +
  theme(plot.title = element_text(size = 15),
        axis.title.x = element_text(size = 14),
        axis.title.y = element_text(size = 14),
        legend.position = "none")+
  scale_x_discrete(limits = top_five_total$common_name)
```

Top five birds with most observations per each year
```{r}
top_five <- total_data %>%
  filter(year_y %in% c(2019, 2020, 2021)) %>% 
  group_by(year_y, common_name) %>%
  summarise(count = n()) %>%
  arrange(desc(count)) %>%
  slice_head(n = 5)
top_five
```

Yellow-rumped Warbler 2019 (to readjust battle)
```{r}
total_data %>% 
  filter(year_y=="2019", common_name=="Yellow-rumped Warbler") %>%
  count()
```

White-crowned Sparrow 2019 & 2019+2020 (to readjust battle)

```{r}
total_data %>% 
  filter(year_y=="2019", common_name=="White-crowned Sparrow") %>%
  count()
```

```{r}
total_data %>% 
  filter(year_y %in% c(2019, 2020), common_name=="White-crowned Sparrow") %>%
  count()
```

Finding totals as years stack
```{r}
top_observations_19 <- total_data %>% 
  filter(year_y=="2019") %>% 
  group_by(common_name) %>% 
  summarise(count = n()) %>%
  arrange(desc(count)) %>%
  slice_head(n = 5)
top_observations_19
```

```{r}
top_observations_19 %>% 
  ggplot(aes(x = common_name, y = count, fill = common_name)) +
  geom_bar(stat = "identity") +
  coord_flip()+
  scale_fill_brewer(palette = "Set3")+
  labs(title = "Top Bird Observations 2019",
       x = "Common Name",
       y = "Count of Observations",
       fill="Common Name") +
  theme(plot.title = element_text(size = 15),
        axis.title.x = element_text(size = 14),
        axis.title.y = element_text(size = 14),
        legend.position = "none")+
  scale_x_discrete(limits = top_observations_19$common_name)
```

```{r}
top_observations_19_20 <- total_data %>% 
  filter(year_y %in% c(2019, 2020)) %>% 
  group_by(common_name) %>% 
  summarise(count = n()) %>%
  arrange(desc(count)) %>%
  slice_head(n = 5)
top_observations_19_20
```
```{r}
top_observations_19_20 %>% 
  ggplot(aes(x = common_name, y = count, fill = common_name)) +
  geom_bar(stat = "identity") +
  coord_flip()+
  scale_fill_brewer(palette = "Set3")+
  labs(title = "Top Bird Observations 2019 & 2020",
       x = "Common Name",
       y = "Count of Observations",
       fill="Common Name") +
  theme(plot.title = element_text(size = 15),
        axis.title.x = element_text(size = 14),
        axis.title.y = element_text(size = 14),
        legend.position = "none")+
  scale_x_discrete(limits = top_observations_19_20$common_name)
```

```{r}
top_observations_19_20_21 <- total_data %>% 
  filter(year_y %in% c(2019, 2020, 2021)) %>% 
  group_by(common_name) %>% 
  summarise(count = n()) %>%
  arrange(desc(count)) %>%
  slice_head(n = 5)
top_observations_19_20_21
```
```{r}
top_observations_19_20_21 %>% 
  ggplot(aes(x = common_name, y = count, fill = common_name)) +
  geom_bar(stat = "identity") +
  coord_flip()+
  scale_fill_brewer(palette = "Set3")+
  labs(title = "Top Bird Observations 2019, 2020, & 2021 Combined",
       x = "Common Name",
       y = "Count of Observations",
       fill="Common Name") +
  theme(plot.title = element_text(size = 15),
        axis.title.x = element_text(size = 14),
        axis.title.y = element_text(size = 14),
        legend.position = "none")+
  scale_x_discrete(limits = top_observations_19_20_21$common_name)
```


Top 5 observations in separate years organized from least to most
```{r}
top_five %>%
  ggplot(aes(x = year_y, y = count, fill = common_name)) +
  geom_bar(stat = "identity", position = "dodge") +
  coord_flip()+
  scale_fill_brewer(palette = "Set3")+
  labs(title = "Top Five Birds with Most Observations",
       x = "Year",
       y = "Count of Observations",
       fill="Common Name") +
  theme(plot.title = element_text(size = 15),
        axis.title.x = element_text(size = 14),
        axis.title.y = element_text(size = 14))
```

Top 5 Birds By Season

```{r}
seasonal_birds <- total_data %>% 
  filter(!is.na(common_name)) %>% 
  group_by(season, common_name) %>% 
  summarise(count = n()) %>%
  arrange(desc(count)) %>%
  slice_head(n = 5)
seasonal_birds
```
```{r}
seasonal_birds %>%
  ggplot(aes(x = season, y = count, fill = common_name)) +
  geom_bar(stat = "identity", position = "dodge") +
  coord_flip()+
  scale_fill_brewer(palette = "Set3")+
  labs(title = "Top Five Birds with Most Observations by Season",
       x = "Season",
       y = "Count of Observations",
       fill="Common Name") +
  theme(plot.title = element_text(size = 15),
        axis.title.x = element_text(size = 14),
        axis.title.y = element_text(size = 14))+
  scale_x_discrete(limits = c("Winter", "Spring", "Summer", "Fall"))
```

**Shiny App**

Questions - 

**What is your favourite season? (Winter, Spring, Summer, Fall)
**Do you prefer when people go the extra mile for you, would you rather meet them halfway, or do you like when others do all the work? (effort_distance_km_y >= 1, effort_distance_km<= 1, effort_distance_km == NA (or 0))
**Do you prefer to be around a lot of people or do you like to keep to yourself? (observation_count >= 5)
**Where is your favourite place to hang out in Davis? (different hotspots - not sure this will work though, need distinctly varying birds per hotspot)
**Do you like the spotlight or prefer to remain in the shadows? (number_observers_y >= 5)





